# Package Vignette {#vignette}

## Forestry Data

```{r, message=F, warning=F, echo= F}
library(here)
library(tidyverse)
library(gt)
data <- readRDS(here("data", "wa_plots_public.rds"))
```

In order to show off the functionality of the package that we wrote, we'll apply it to the Forestry data that we introduced all the way back in Chapter \@ref(intro-section). Recall that this forestry data is both zero-inflated and clustered making it a perfect setting for us to employ our models.

The particular subset of Forestry data that we'll be looking at comes from the State of Washington and contains `r nrow(data)` observations across 49 Eco-Subsections (which define our clusters). The summary statistics of the number of observations per group are as follows:

```{r, message=F, warning=F, echo=F}
summary(data %>% count(ECOSUBCD) %>% pull(n)) 
```





We start by installing the developmental version of the package and loading it into our R session

```{r, message=F, warning=F}
devtools::install_github("joshyam-k/zinf")
library(zinf)
```

